```{r}
#| label: setup-corstr
#| include: false
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, fig.width=6.5, fig.height=2.5)
```

# Correlation Structures

This section is an appendix to the GEE chapter, to illustrate and
discuss the different correlation structures that we can specify when
fitting a GEE.

### Variance/Covariance or Correlation?

Statisticians often talk about a model's **variance-covariance** matrix
(which gives residual variance on the diagonal and covariance between
residuals in its off-diagonal elements). Here we will consider
**correlation** matrices instead. This will give us a simplified way of
looking at similar ideas. In a correlation matrix, the diagonal entries
will be all 1s (the correlation of something with itself is 1), and
off-diagonal elements will show correlation between residuals.

### Example Case

Let's consider an example for a dataset with 9 observations; three
observations for each of three individuals. *(Side note: The groupings
causing non-independence in the residuals don't have to be "individuals"
-- it could be some other relationship, like being from the same school
or town or country or ethnic group -- any single categorical variable
that defines the groups of interest and induces non-independence withing
the groups, could work. Here we are just calling them "individuals" for
convenience, and because when time-series data are collected on multiple
individuals, this kind of model often works well.)*

Each of the correlation matrices below has one row and one column for
each observation.

The *diagonal* entries will all, always, be 1.

The *off-diagonal* entries indicate how the different residuals depend
on each other -- in other words, they describe mathematically exactly
*how* each residual is correlated with the others.

We will use the letter $\rho$ for correlations.

### Independence

If the residuals are all independent of each other, then all the
correlations between them will be zero:

$$
\begin{vmatrix}
1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
\end{vmatrix}
$$

This corresponds to all of the models we have considered previously in
class (which all have a condition that residuals should be independent).

### ACF Example

The residual ACF from a model fitted to data that perfectly embody this
structure with 5 observations per individual might look like:

```{r, echo=FALSE}
d <- data.frame(r = as.numeric(arima.sim(model = list(), n=5)))
while(nrow(d) < 100){
  a <- data.frame(r = as.numeric(arima.sim(model = list(), n=5)))
  d <- bind_rows(d,a)
}
acf(d$r, main='Residual ACF')

```

### Exchangeable = Block Diagonal

In an exchangeable or block diagonal structure, all residuals for one
individual are equally correlated (correlation = $\rho$). All residuals
from different individuals are independent (correlation = 1).

$$
\begin{vmatrix}
1 & \rho & \rho & 0 & 0 & 0 & 0 & 0 & 0\\
\rho & 1 & \rho & 0 & 0 & 0 & 0 & 0 & 0\\
\rho & \rho & 1 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & \rho &\rho & 0 & 0 & 0\\
0 & 0 & 0 & \rho & 1 & \rho & 0 & 0 & 0\\
0 & 0 & 0 & \rho & \rho & 1 & 0& 0& 0\\
0 & 0 & 0 & 0 & 0 & 0 & 1 & \rho & \rho\\
0 & 0 & 0 & 0 & 0 & 0 & \rho & 1 & \rho\\
0 & 0 & 0 & 0 & 0 & 0 & \rho & \rho & 1\\
\end{vmatrix}
$$

#### ACF Example

The residual ACF from a model fitted to data that perfectly embody this
structure would show non-independence in the residual ACF, out to about
the number of lags that there are observations per individual, but the
autocorrelation coefficients would not decline at exactly the rate
predicted by an AR(1) process (see below).

### AR1 (first-order autoregressive process)

In this structure, we again assume that residuals are uncorrelated
(independent) between individuals.

Within an individual, we assume measurements that are closer to one
another are more correlated. More precisely, we say that observations 1
lag apart have correlation $\rho$; observations 2 lags apart have
correlation $\rho^2$; three lags apart $\rho^3$, and so on.

To illustrate this, I show below a correlation structure for a model
with 10 observations: 5 observations from each of 2 individuals.

$$
\begin{vmatrix}
1 & \rho & \rho^2 & \rho^3 & \rho^4 & 0 & 0 & 0 & 0 & 0\\
\rho & 1 & \rho &\rho^2 & \rho^3 & 0 & 0 & 0 & 0& 0\\
\rho^2 & \rho & 1 & \rho & \rho^2 & 0 & 0 & 0 & 0& 0\\
\rho^3 & \rho^2 & \rho & 1 & \rho & 0 & 0 & 0 & 0& 0\\
\rho^4 & \rho^3 & \rho^2 & \rho & 1 & 0 & 0 & 0 & 0& 0\\
0 & 0 & 0 & 0 & 0 & 1 & \rho & \rho^2 & \rho^3 & \rho^4\\
0 & 0 & 0 & 0 & 0 & \rho & 1 & \rho &\rho^2 & \rho^3\\
0 & 0 & 0 & 0 & 0 & \rho^2 & \rho & 1 & \rho & \rho^2 \\
0 & 0 & 0 & 0 & 0 & \rho^3 & \rho^2 & \rho & 1 & \rho\\
0 & 0 & 0 & 0 & 0 & \rho^4 & \rho^3 & \rho^2 & \rho & 1\\
\end{vmatrix}
$$

#### ACF Example

The residual ACF from a model fitted to data that perfectly embody this
structure with $\rho=0.99$ and 5 observations per individual, might look
like:

```{r, echo=FALSE}
d <- data.frame(r = as.numeric(arima.sim(model = list(ar = 0.99), n=5)))
while(nrow(d) < 100){
  a <- data.frame(r = as.numeric(arima.sim(model = list(ar = 0.99), n=5)))
  d <- bind_rows(d,a)
}
acf(d$r, main='Residual ACF')
```

### Unstructured

This one is not of practical use because it is so hard to estimate.

$$
\begin{vmatrix}
1 & \rho_{1,2} & \rho_{1,3} & 0 & 0 & 0 & 0 & 0 & 0\\
\rho_{1,2} & 1 & \rho_{2,3} & 0 & 0 & 0 & 0 & 0 & 0\\
\rho_{1,3} & \rho_{2,3} & 1 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & \rho_{1,2} & \rho_{1,3} & 0 & 0 & 0\\
0 & 0 & 0 & \rho_{1,2} & 1 & \rho_{2,3}& 0 & 0 & 0\\
0 & 0 & 0 & \rho_{1,3} & \rho_{2,3} & 1&0 & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 1 & \rho_{1,2} & \rho_{1,3}\\
0 & 0 & 0 & 0 & 0 & 0 & \rho_{1,2} & 1 & \rho_{2,3}\\
0 & 0 & 0 & 0 & 0 & 0 & \rho_{1,3} & \rho_{2,3} & 1\\
\end{vmatrix}
$$
