```{r}
#| label: setup-gam
#| include: false
library(tidyverse)
library(mosaic) 
library(ggformula)
theme_set(theme_bw(base_size=14))
knitr::opts_chunk$set(
  tidy=FALSE, echo=TRUE, 
  size="small", message=FALSE, results='markup',
  fig.width=6, fig.height=3.5, cache=FALSE)
```

# GAMs: Generalized Additive Models

So far, we have learned ways to model continuous, logical, and count
response variables as functions of quantitative and categorical
predictors. We started with linear models - where both the response and
predictor variables are quantitative and the relationship between them
is *linear*. What about nonlinear relationships?

So far, we have considered...

-   **Categorical predictor variables**. Making use of indicator
    variables for (all but one of the) categories, we can model a
    situation where each value of the predictor variable has a different
    effect on the response. But...
    -   How many categories?
    -   What about periodicity?
-   **GLMs**. In logistic, Poisson, etc. regression, the action of the
    link function results in a relationship between the predictors and
    the response variable that is linear on the scale of the link
    function ( = scale of the RHS of the equation), but non-linear on
    the scale of the response variables (LHS). But...
    -   Nonlinear, but **monotonic**

## Non-linear, non-monotonic relationships

It's not true that all interesting predictor-response relationships are
linear or monotonic. One example is in some data on bat migration: how
does the probability of bats leaving on their migratory journey depend
on air humidity?

```{r}
bats <- read.csv('https://ndownloader.figshare.com/files/9348010')
```

```{r, include = FALSE}
    # old url: 'http://rsbl.royalsocietypublishing.org/highwire/filestream/33982/field_highwire_adjunct_files/2/rsbl20170395supp3.csv')
```

```{r,fig.width=6.5, fig.height=2, fig.show = 'hold', echo = FALSE}
gf_boxplot(relativehumid ~ factor(migration), data = bats)

bats <- bats |>
  mutate(humid_bins = cut(bats$relativehumid, breaks=7))

bats2 <- bats |>
   group_by(humid_bins) |>
  summarize(migration_prop = prop(~migration == 1)) |>
  ungroup()

gf_point(migration_prop ~ humid_bins, data = bats2) |> 
  gf_labs(x='Relative Humidity', y='Proportion Migrating')
```

Another dataset (our example for the day) -- ozone levels as a function
of temperature, day, and humidity:

```{r, fig.width=2.2, fig.height=2, fig.show='hold'}
ozone <- read_csv('https://raw.githubusercontent.com/selva86/datasets/master/ozone.csv') |>
  mutate(Date = lubridate::mdy(paste(Month, Day_of_month, '2018')),
         Day = lubridate::yday(Date))
gf_point(ozone_reading ~ Temperature_Sandburg, 
         data=ozone, alpha=0.4, ylab = 'Ozone')
gf_point(ozone_reading ~ Day, 
         data=ozone, alpha=0.4, ylab = 'Ozone')
gf_point(ozone_reading ~ Humidity, 
         data=ozone, alpha=0.4, ylab = 'Ozone')
```

## Smooth terms

We can fit a model where the relationship between the response and the
predictor is a \`\`smooth" -- no linearity or monotonicity requirement.

### Basis functions

-   A smooth term is constructed as the sum of several parts, or *basis
    functions*. Each of these functions has a relatively simple shape,
    but scaled and added together, they can produce nearly any
    \`\`wiggly" shape.
-   Increasing the dimension of the basis (more functions added
    together) can allow more wiggliness.
-   Goal: allow enough wiggliness to fit the data well, without
    *overfitting* (smooth goes through every point in the data, or
    follows \`\`trends" that are spurious)

We will fit smooth models to data using the function `gam()` from the
package `mgcv`. It includes many options for basis functions (types of
smooths) - see `?mgcv::gam` or
\[https://rsconnect.calvin.edu:3939/content/28/\] for details.

## Fitting GAMs

An excellent resource: <https://converged.yt/mgcv-workshop/>.

### Choosing model formulation

Which terms should be modelled as smooth terms? Explore the data!

-   Pros:
-   Cons:

### Model formula

Let's fit a simple GAM for the ozone data as a function of radiation,
temperature and wind. Note the `s()` function for specifying a smooth,
which takes as input:

-   a variable name (or more than one, for advanced users)
-   `k`
-   `bs`

How do we choose? For some exploration, see:
\[https://rsconnect.calvin.edu:3939/content/28/\].

We can also fit the model and smooths by different methods and with
options:

-   `method = 'GCV.Cp'`
-   `method = 'REML'`
-   `method = 'ML'`
-   `select = TRUE` (or `FALSE`)

```{r, fig.show='hold', fig.width=2.2, fig.height=2.25, results='markup'}
library(mgcv)
oz.gam <- gam(ozone_reading ~ s(Day, k = 7, bs = 'cc') +
                s(Wind_speed, k = 5, bs = 'tp') +
                s(Temperature_Sandburg, 
                  k = 5, bs = 'tp'),
              data = ozone,
              method = 'ML', 
              select = TRUE)
summary(oz.gam)
```

## Model Assessment

In addition to what you already know (...which all still holds, except
linearity expectation!) `mgcv` has some nice model checking functions
built in.

```{r, fig.show='hold', fig.width=5.5, fig.height=3.5, results='markup'}
par(mar=c(4,4,2,2))
gam.check(oz.gam)
```

### Concurvity

Like collinearity and multicollinearity, but for smooths...values of 0
indicate no problem, and 1 a huge problem (total lack of identifiability
-- same information in multiple predictors).

Overall, does the model have problems with concurvity?

```{r}
concurvity(oz.gam, full=TRUE)
```

Or alternatively, which specific pairs of terms cause problems?

```{r}
concurvity(oz.gam, full=FALSE)
```

## Model Selection

### Shrinkage and Penalties

With GAMs, in a sense, some model selection is (or can be) done during
model fitting - what smooth is best? Or is the relationship a line? A
flat line? Using *shrinkage* basis or including `select=TRUE` allows for
this.

### P-value selection

Cautions: **p-values are approximate!** Successfulness of the procedure
best when fitting method is: ML (1st choice), REML (2nd choice).

```{r, results = 'hide'}
anova(oz.gam)
```

Interpretation as usual. Note that `anova()` (not `Anova()`) works
here - especially for GAMs, it does *not* do sequential tests; and
`Anova()` doesn't handle GAMs.

### Information criteria

-   Conditional/Approximate - bias
-   Fitting method:
    -   REML-based IC scores can be used to compare models with
        different *random effects* but **not** different predictors. (IF
        `select = TRUE` and using a shrinkage basis.)
    -   ML-based IC scores can be used to compare models with different
        fixed effects (regular predictors) and different `family`, but
        **not** different random effects

```{r}
library(MuMIn)
oz.ml <- update(oz.gam, method='ML', na.action='na.fail')
head(dredge(oz.ml, rank='AIC'),2)
```
