```{r}
#| label: setup-gee
#| include: false
library(mosaic)   
library(ggformula)
library(tidyverse)
library(s245)
library(MuMIn)
theme_set(theme_bw(base_size=18))
knitr::opts_chunk$set(
  tidy=FALSE,    
  size="footnotesize",   
  fig.width=3.5, fig.height=2, fig.show='hold')
```

# GEEs

So far, we looked at random effects as one way to account for
non-independence of residuals due to a variable that we don't want to
include as a regular predictor. Another option for this kind of
situation is to use Generalized Estimating Equations (GEEs).

## Data Source

The dataset used here is industry data from a skin care company. It
contains data from experiments with 20 subjects. Each person tested 6
different skin moisturizers, and the hydration level of their skin was
measured every 2 hours for 24 hours following application of each
product. The variables are:

-   `Subjects` Numeric code identifying the person
-   `CorneoDiff` The hydration CorneoDiff
-   `Time` Time in hours since product application
-   `Product` Which product was used

The data file can be accessed online at:

<http://sldr.netlify.com/data/hydrationData.csv>

```{r, hyddat, echo=FALSE, fig.width=6.5, fig.height=2.5}
hyd <- read.csv('http://sldr.netlify.com/data/hydrationData.csv')
hyd <- hyd |> mutate(Subjects = factor(Subjects),
                      Product = factor(Product))
hyd.sm <- hyd |> dplyr::filter(Subjects %in% c('1', '2', '3','4','5', '6'))
```

## Data Exploration

We would like to model the hydration, `CorneoDiff`, over time and as a
function of product.

```{r, fig.width=7, fig.height=3.5, warning=FALSE}
gf_point(CorneoDiff ~ Time | Product, data=hyd.sm) |>
  gf_lims(x=c(0,24)) 
```

```{r, fig.width=7, fig.height=3.5, warning=FALSE}
gf_point(CorneoDiff ~ Time | Subjects, color=~Product, data=hyd.sm) |>
  gf_line(CorneoDiff ~ Time | Subjects, color=~Product, data=hyd.sm) |>
  gf_lims(x=c(0,24)) 
```

## Linear Regression

We could try just fitting a linear regression. What do you expect?

```{r}
lm1 <- lm(CorneoDiff ~ Time + Product, data=hyd.sm)
summary(lm1)
```

```{r, fig.width=7, fig.height=3.5, warning=FALSE}
gf_point(CorneoDiff ~ Time | Subjects, color=~Product, data=hyd.sm) |>
  gf_line(CorneoDiff ~ Time | Subjects, color=~Product, data=hyd.sm) |>
  gf_lims(x=c(0,24))  |>
  gf_abline(intercept=7.86, slope=0.05608)
```

## Model Assessment

For the linear regression:

```{r, fig.width=6, fig.height=3.5}
acf(resid(lm1))
```

```{r, fig.width=6, fig.height=3.5}
hyd.sm <- arrange(hyd.sm,Time, Subjects)
lm1 <- lm(CorneoDiff ~ Time + Product, data=hyd.sm)
acf(resid(lm1))
```

*As we expected, things do not look good...*

## Linear Regression

We tried a linear regression and encountered two problems:

-   The residuals are not independent. There seems to be correlation
    over time within subjects.
-   We can't account for inter-person differences unless we include
    person as a predictor, but we don't want to do that, because if we
    do we can not make predictions from the fitted model without
    specifying which of these exact people we want to predict for.
    That's not ideal - we want predictions for all people, or at least
    averaged over all people.

## Generalized Estimating Equations (GEEs)

A potential solution we will invesitgate today is to use a generalized
estimating equation (GEE) instead of a GLM. GEEs:

-   Are a \`\`PA" model:
-   Work by changing... </br> </br> </br> </br>

What residual correlation structures can be accomodated in this
framework?

-   Independence (`corstr=`independence'\`)
-   Exchangeable = Block Diagonal (`corstr=`exchangeable'\`)
-   AR1 (first-order auto-regressive) (`corstr=`ar1'\`)
-   Unstructured (CAUTION!) (`corstr=`unstructured'\`)

### Fitting GEEs with different correlation structures

```{r,}
library(geepack)
hyd <- arrange(hyd, Subjects, Time)
lm1 <- lm(CorneoDiff ~ Time + Product, data=hyd)
summary(lm1)
gee.ind <- geeglm(CorneoDiff ~ Time + Product, data=hyd, 
                  id = Subjects, corstr='independence')
summary(gee.ind)
gee.ar1 <- geeglm(CorneoDiff ~ Time + Product, data=hyd,
                  id = Subjects, corstr='ar1')
summary(gee.ar1)
gee.exch <- geeglm(CorneoDiff ~ Time + Product, data=hyd, 
                   id = Subjects, corstr='exchangeable')
```

What is **the same** (or similar) and what is very **different** between
the models?

### Comparing different correlation structures

We can use a specific variance of QIC, $QIC_{R}$, to compare models with
different correlation structures:

```{r,}
library(MuMIn)
# QIC(gee.ind, gee.exch, gee.ar1, typeR=TRUE)
```

How can we interpret this result?

## GEE model assessment

Model assessment for a GEE is mostly the same as for the corresponding
linear regression or GLM (Poisson, Logistic, etc.)

We were using GEEs to try to correct for issues with non-independent
residuals. How does the residual plot change for a GEE relative to the
corresponding (g)lm? *Does* it change? *Should* it?

```{r,fig.width=3.5, fig.height=2.8}
acf(resid(lm1), main='LM ACF')
acf(resid(gee.ind), main='GEE ACF')
```

**What is going on here?**

## Model Selection - Which variables?

We can use another variant of the QIC to do model selection to determine
which variables are important to retain in a GEE model.

```{r}
gee.ind <- update(gee.ind, na.action='na.fail')
dredge(gee.ind, rank='QIC', typeR=FALSE)
```

How would you interpret these results and present them to the cosmetics
company that collected the data?

## Prediction Plots

As for models we studied previously, we can make prediction plots to
visualize the relationships a model specifies between the predictor and
response variables.

However, we can not use `predict()` to get model predictions **with
standard errors** from a GEE.

`pred_plot()` works, though; for example:

```{r}
s245::pred_plot(gee.ind, 'Product') |>
  gf_labs(y = 'CorneoDiff')
```

Once again we're grateful for the parametric bootstrap! (This time,
`pred_plot()` is silently doing the work for us.)
