```{r}
#| label: setup-interactions
#| include: false

library(tidyverse)   
library(ggformula)
library(mosaic)
library(s245)

theme_set(theme_bw(base_size=12))     

knitr::opts_chunk$set(
  echo = TRUE,      
  tidy = FALSE,     
  size = "small",   
  message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 2.1, fig.align = 'center') 

teach_beauty <- read.csv('http://www.stat.columbia.edu/~gelman/arm/examples/beauty/ProfEvaltnsBeautyPublic.csv') |>
  rename(beauty = btystdave,
         eval = courseevaluation) |>
  mutate(race_eth = ifelse(minority == 0, 'White/Caucasian', 'Other'),
         language = ifelse(nonenglish == 1, 'English', 'Other Language'),
         formal = ifelse(formal == 1, 'Formal Dress', 'Informal Dress'),
         female = ifelse(female == 1, 'female', 'not female')) |>
  dplyr::select(beauty, eval, race_eth, language, age, students, formal, profnumber, female)
```

# Interactions

Two predictors **interact** when you need to know values of *both* in
order to make an accurate prediction of the response variable value.

Predictors can interact in *any* type of regression model (so this
chapter could really be placed almost anywhere).

## Example: Quantitative-Categorical Interaction

```{r}
gf_point(eval ~ beauty, color = ~female, shape = ~female, data = teach_beauty) |>
  gf_lm()
```

`Eval` may go up as `beauty` increases, but the *slope* of the
relationship is *different* for females and non-females. This is an
**interaction** between `beauty` and `female`.

## Categorical-Categorical Interaction Example

```{r}
gf_boxplot(eval ~ formal | female, data = teach_beauty)
```

Perhaps *Informal Dress* affects `eval` scores, but really only for
non-females -- for females, `formal` dress doesn't make a difference
either way.

The effect of `formal` dress is *different* depending on the value of
`female`. This is an interaction between `formal` and `female`.

## Quant-Quant interactions?

Yes, these are possible, but very hard to visualize and conceptualize.
Basically, it would mean that the slope of the line for one predictor
changes gradually as the value of a second variable changes.

## R code

If you want to include an interaction term in a model in R, use a `*`
rather than a `+` between the predictors that (may) interact. For
example, based on our exploration above, we might try:

```{r}
beauty_mod <- lm(eval ~ beauty*female +
                   formal*female, 
                 data = teach_beauty, 
                 na.action = 'na.fail')
summary(beauty_mod)
```

Notice the additional indicator variables in the coefficient table/model
equation. Now we need to *adjust* the effects of the `beauty` predictor
depending on the values of `formal` and `female`, which interact with
it.

We can use IC-based model selection to determine whether including these
interactions in a model is important or not.

```{r}
library(MuMIn)
dredge(beauty_mod, rank = 'AIC')
```

In the case of the particular model we fitted, the "best" model starting
from this full model is actually one *without* interactions. If you want
to explore the dataset further, you will find that actually a model
where `age`, `beauty` AND `female` interact fits much better...

## Cautionary note

If you include an interaction in a regression model, you **must** also
include the corresponding "fixed effects" -- this means if you have an
indicator variable/slope term for an interaction in your model, you must
also have the indicator variables/slopes corresponding to the individual
predictors. Our fitting functions (`lm()`, `glm()`, `glmmTMB()`, etc.)
are smart enough to ensure this for you. So is `dredge()`. (It would
take effort to mess this up in R.)
