```{r}
#| label: setup-zeroinfl
#| include: false
library(mosaic)   
library(ggformula)
library(pscl)

theme_set(theme_bw(base_size=18)) 
knitr::opts_chunk$set(
  tidy=FALSE,     
  size="small", message=FALSE, warning=FALSE,
  fig.width = 3.5, fig.height = 2.2, fig.align = 'center', fig.show = 'hold')   
```

# Zero-Inflation

Zero-inflated data are count data that:

-   Have a lot of zeros; **and**
-   It may not be appropriate to explain the extra zeros via a model for
    count data with overdispersion (NB or quasi-Poisson) because:
    -   There is too much overdispersion to account for this way
    -   The zeros in the data are of two classes: *true* zeros (definite
        absence) and *other* zeros (might have been a positive count but
        just happened to be a 0 this time).

## Reference material

On the topic of zero-inflated data (as on many other topics), your book
is ...terse! I have several copies of [Mixed models and extensions for
ecology in
R](https://www.highstat.com/index.php/our-books?view=article&id=17&catid=18 "link the the Zuur et al "Penguin book"")
if you want to borrow -- You may want to check out Chapter 11 in that
book.

## Data for Example

The state wildlife biologists want to model how many fish are being
caught by fishermen at a state park. Visitors are asked how long they
stayed, how many people were in the group, were there children in the
group and how many fish were caught. Some visitors do not fish, but
there is no data on whether a person fished or not. Some visitors who
did fish did not catch any fish so there are excess zeros in the data
because of the people that did not fish.

We have data on 250 groups that went to a park. Each group was
questioned about how many fish they caught (count), how many children
were in the group (child), how many people were in the group (persons),
and whether or not they brought a camper to the park (camper).

In addition to predicting the number of fish caught, there is interest
in predicting the existence of excess zeros, i.e., the probability that
a group caught zero fish. We will use the variables child, persons, and
camper in our model. Let's look at the data.

```{r}
zinb <- read.csv("https://stats.idre.ucla.edu/stat/data/fish.csv")
head(zinb)

zinb <- zinb |> mutate(nofish=factor(nofish),
                        livebait=factor(livebait),
                        livebait = ifelse(livebait==0, 'No', 'Yes'),
                        camper=factor(camper),
                        camper = ifelse(camper ==0, 'No', 'Yes'),
                        child_cat = factor(child))
head(zinb)
```

## Visualization?

In class, we created several figures for data exploration. They are not
printed here.

```{r, include = FALSE, results = 'hide', fig.show = 'hide'}
gf_histogram(~count, data=zinb) |>
  gf_labs(y='Number of Observations', x='Fish Caught')

gf_boxplot(count ~  factor(child), data=zinb)
gf_boxplot(log(count) ~ livebait, data=zinb)
gf_boxplot(log(count) ~ camper, data=zinb)
```

One idea was to make a pairs plot to look at possible collinearity. Good
idea! But all our predictors are either yes/no or have very small ranges
of integers as their values, so it doesn't prove very informative with
our data - feel free to try for yourself to illustrate:

```{r, eval=FALSE, echo=TRUE}
library(GGally)
ggpairs(zinb[, c('livebait', 'camper', 'persons', 'child')])
```

Another idea would be to plot response as a function of possible
predictors, to check the linearity conditions of the parts of the model.
This could be a good idea, but on the other hand we don't have any
predictors that are unequivocally numerical. (If we had one, we could
plot log(count) as a function of the predictor.)

## Collinearity/Multicollinearity?

We might want to check VIFs, but we can't compute them directly for a
ZIP model (fitted later) using `vif()` from the `car` package.

However, note: VIFs really depend mainly on the covariates, not the
model fitted. We can check them effectively by fitting the Poisson model
(and the logistic regression, if we had more than one predictor in that
part of the model) and checking **its** VIFs:

```{r}
P <- glm(count ~ camper + child + persons, data=zinb, family=poisson)
library(car)
vif(P)
```

No evidence of any problem there!

## Fitting models

### Zero-inflated Poisson

```{r}
library(pscl)
ZIP <- zeroinfl(count ~ child + camper + persons | 1, data = zinb,
                dist = 'poisson')
summary(ZIP)
```

### Zero-inflated NB: one way

We can also use this function to fit a NB model. *BUT BEWARE:* this is a
different parameterization of the NB model than the NB1 and NB2 models
we use in `glmmTMB()` and may be closer to type 1 or type 2 depending on
the dataset. You could compare all three NB versions via IC.

```{r}
ZINBx <- zeroinfl(count ~ child + camper + persons | 1, data = zinb,
                  dist='negbin')
summary(ZINBx)
```

#### What do the `summary()` outputs mean?

The second part of the output (labelled "Zero Inflation model
coefficients") is from a logistic regression with one predictor
(persons) to predict the probability of an observation being a "true
zero" - a group that was not in the park to fish. (We can't really tell
from thsi output whether being there to fish, or not, is considered a
true zero by the model...I figured it out by reading the help file for
zeroinfl and looking at prediction plots below). The first part of the
out put (labelled "Count model coefficients") is a Poisson regression
fitted to the count data for the proportion of the groups in the data
set who were *not* true zeros (and were thus actually there to fish).
Both parts of the model can have predictors included; to have no
predictors in the logistic regression part, you would use a formula like
**y \~ x1 + x2 ... + xn \| 1**, and to include the *same* predictors for
both the Poisson and logistic parts, you exclude the **\|...** part, for
example \*\* y \~ x1+x2\*\* is the same as **y \~ x1 + x2 \| x1 + x2**.

### Zero-inflated negative binomial (other way)

```{r}
library(glmmTMB)
ZINB1 <- glmmTMB(count ~ child + camper + persons, ziformula = ~1, data=zinb, family=nbinom1(link='log'))
ZINB2 <- glmmTMB(count ~ child + camper + persons, ziformula = ~1, data=zinb, family=nbinom2(link='log'))
summary(ZINB1)
```

Note the difference from the `zeroinfl()` model formula and function
call. There is an separate input, `ziformula`, to add zero inflation to
a `glmmTMB()` model. (The default is no zero inflation, if you do not
include a `ziformula`). The zero-inflation model always uses a logit
link.

### Tweedie Model

Another option for this type of data might be to try a Tweedie model,
because it also has a probability mass at 0 (in other words, allows for
zero inflation). The Tweedie distributions we are considering are not
exactly perfect for count data since the positive part of the
distribution is like a gamma distribution (continuous, not counts), but
the shape and support of the distribution are pretty close to our data.
Let's try:

```{r}
ZIT <- glmmTMB(count ~ child + camper + persons, data=zinb, family=tweedie(link = 'log'))
summary(ZIT)
```

## Model Assessment

We can do model assessment "like" we do for other count data
models...more or less. (It is much harder to get something like Pearson
residuals for these models, though...) *Please stay tuned!*

## Model Selection?

One thing to understand is: is the ZIP model a better fit to the data
than just a Poisson model? We can think of ZI as a different way of
modelling overdispersion, an alternative to NB or quasi-Poisson models
when we know there are true zeros in the data set.

```{r}
P <- glm(count ~ camper + child + persons, data=zinb, family=poisson)
NB1 <- glmmTMB(count ~ camper + child + persons, data=zinb,
              family=nbinom1(link='log'))
NB2 <- glmmTMB(count ~ camper + child + persons, data=zinb,
              family=nbinom2(link='log'))
AIC(ZIP,P, NB1, NB2)
```

Well, yes, the ZIP model is much better than the Poisson! But, the NB1
and NB2 models are *much* better than both.

Among the different families and types of models that we have tried,
which seems to be best for this data set?

```{r}
BIC(P, ZIP, ZINB1, ZINB2, ZINBx, ZIT)
```

For these data, the zero-inflated models are very similar to the other
NB models, in terms of the AIC. If we used BIC instead, the non-zero
inflated models would look quite a bit better! The improvement in
goodness of fit with the zero-inflation is not really enough to make us
prefer these ZI models to simpler models that also allow overdispersion,
like the NB models.

The NB models and the ZI models (with whatever "family" you choose) are
two *different* ways of dealing with overdispersion caused by an
abundance of zeros in the data. I'd recommend choosing ZI models if:

-   There are *true zeros* and *other zeros* in your data, and you want
    to model this explicitly; or if
-   The ZI models fit the data much better than a simpler non-ZI NB (or
    perhaps Tweedie) model.

The zero-inflated negative binomial models seem to be best, among the ZI
models. However, there is still a lot more work that could be done here!
For example...

### Zero inflation covariates

what if the Logistic regression part of the model *also* depends on
covariates? That seems reasonable -- maybe the number of people present,
whether they are camping, and the number of kids present would help you
to predict whether a group was not going to go fishing at all. To check
that, we would need to change the `ziformula` (in NB models) or the part
of the formula after the $\vert$ in ZIP models. It is possible this
change **might** make the ZI models competitive with, or better than,
the non-ZI NB models in IC terms.

### Interaction terms

For this dataset, it may also make sense to consider some more
predictors, and interaction terms. For example, maybe live bait use is a
good predictor, and maybe the relationship between live bait use and
number of fish caught changes depending on the number of children
present -- for example, we might try something like:

```{r}
ZINB1b <- glmmTMB(count ~ camper + child*livebait + persons, ziformula= ~1, 
                  data=zinb, family=nbinom1(link='log'))
summary(ZINB1b)
AIC(ZINB1, ZINB1b)
```

Hmmm...it looks like we made an improvement. But do you think the
improvement is because of the interaction term?

Looking at the p-values, we might suspect that while `livebait` helps
predict fish caught, there isn't a clear interaction between effects of
live bait use and the number of kids present.

In this case, it is hard to fathom all the model selection options
because there are predictors in *both parts* of the model.

If you continue by hand, one thing to remember is: You can **never**
include an interaction term in a model without also including the
corresponding "main effect" term. For example, you *can not* have an
interaction between `child` and `livebait` without also having both
`child` and `livebait` as individual predictors in the model. (In fact,
I have intentionally not taught you how to fit the "wrong" model...as
long as you use the \* to specify an interaction in your model formula,
you should be fine!)

### Dredge

We *can* also use dredge(), but:

-   It will only put covariates in/out of the model, not compare
    zero-inflated with non-zero-inflated models.
-   It *will* put covariates in and out of both the zero-inflation and
    regular parts of the model.

So to do a comprehensive comparison, you would need to start with
several full models -- one for each "family" you want to try, with and
without ZI, and with all possible predictors included for both the count
and ZI parts of the models -- and compare dredge results for all of
them.

`dredge()` is also nice and smart about interaction terms -- it will
never include an interaction unless the main effect terms are also
present.

```{r, message=FALSE, warning=FALSE}
library(MuMIn)
ZINB1c <- glmmTMB(count ~ child_cat*livebait + camper + persons, 
                  ziformula = ~ child_cat + camper + persons, 
                  data=zinb, family=nbinom1(link='log'), na.action = 'na.fail')
ms <- dredge(ZINB1c, rank='AIC')
head(ms)
```

In this case, which model do you think is best? What else might you try?

## Prediction Plots

We can also use prediction plots as usual. Here, they will help us
clarify what "success" is in the logistic regression part of the model
-- according to the help file for the `zeroinfl()` function, we should
be able to see that a "success" is to be a "true zero".

Let's try an example - making prediction plots for our ZIP model for
different numbers of persons in the group.

First let's check the range of the data for different variables:

```{r}
summary(zinb[,c('camper', 'child', 'persons')])
```

Now, the prediction plots:

```{r}
new_data <- data.frame(persons=c(1,2,3,4),
                       camper='Yes',
                       child=0)
pred <- predict(ZIP, newdata=new_data)
gf_point(pred~persons, data=new_data) |>
  gf_labs(x='People in Group', y='Number of Fish Caught')
```

The number of fish caught goes *up* as the number of people in the group
goes up. But persons was a predictor for the logistic regression
(zero-inflation part) of the model! And its coefficient was estimated to
have a negative value. So according to our model, the more persons in
the group, the less likely you are to be a "true zero" (and the more
likely you are to be in the park to fish). Therefore, the number of fish
your group catches goes up.

However, notice that unfortunately, the `predict()` function for
`zerofinl()` ZIP models does not return standard errors on the fitted
values. Shucks! We would have to compute these ourselves to show
uncertainty on the prediction plot. We could use something called a
parametric or nonparametric bootstrap to do it. We may learn how to do
this later on in the course, but for now we don't have a tool to do it.

If you use one of the `glmmTMB()` models, though, you should be able to
make prediction plots with uncertainty as usual.

## Acknowledgements

Many thanks to: <https://stats.idre.ucla.edu/r/dae/zip/>, a source of
the data set and information used in preparing these notes.
